---
title: "CGM_trad_metrics"
author: "Paul O'Dwyer"
date: "2024-08-05"
output: html_document
---
################################################################################
################################################################################
################################################################################

CGM Metrics Analysis for Multiple Individuals
```{r}
# Load required libraries
library(tidyverse)
library(stats)
library(dplyr)
library(lubridate)
library(iglu)
library(ggplot2)
```

```{r}
# File 2: CGM data (including MAGE)
path <- file.choose()
long_data <- read.csv(path, sep=",", header = TRUE)
```

```{r}
# Filter the data by Center
center_filtered_data <- filter(long_data, Center == "UM")
# Subset the columns
cgm_long_data <- data.frame(
  id = center_filtered_data$Participant,
  period = center_filtered_data$Period,
  gl = center_filtered_data$`Sensor.Glucose..mmol.L`,
  time = center_filtered_data$Timestamp_w_daylight_saving
)
```

```{r}
# Ensure the 'time' column is in POSIXct date-time format
cgm_long_data$time <- as.POSIXct(cgm_long_data$time, format = "%Y-%m-%d %H:%M", tz="UTC")
```

```{r}
cgm_long_id_date_sorted <- cgm_long_data %>% 
  arrange(id, time)
```


```{r}
# Function to calculate metrics for a single day
calculate_daily_metrics <- function(day_data) {
  tryCatch({
    mage_result <- mage(day_data)$MAGE
    peak_glucose <- max(day_data$gl, na.rm = TRUE)
    time_to_peak <- day_data$time[which.max(day_data$gl)]
    
    c(
      mean_glucose = mean(day_data$gl, na.rm = TRUE),
      variance_glucose = var(day_data$gl, na.rm = TRUE),
      mage = mage_result,
      peak_glucose = peak_glucose,
      time_to_peak = as.character(time_to_peak)
    )
  }, error = function(e) {
    c(
      mean_glucose = NA_real_,
      variance_glucose = NA_real_,
      mage = NA_real_,
      peak_glucose = NA_real_,
      time_to_peak = NA_character_
    )
  })
}
```

```{r}
# Function to calculate metrics for each individual
calculate_individual_metrics <- function(individual_data) {
  # Split data by day
  daily_data <- split(individual_data, as.Date(individual_data$time))
  
  # Calculate metrics for each day
  daily_metrics <- lapply(daily_data, calculate_daily_metrics)
  
  # Combine daily metrics
  metrics_df <- do.call(rbind, daily_metrics)
  
  # Add date column
  metrics_df <- as.data.frame(metrics_df)
  metrics_df$date <- names(daily_data)
  
  return(metrics_df)
}
```

# Several issues says some the dates data is not sorted even though it is. 
# Says there are duplicate values for some times, only takes final value, the values generally are very similar so dont really care
# The beginning day is not a full day so could discount this.

```{r}
# Process all individuals
all_metrics <- cgm_long_id_date_sorted %>%
  group_by(id, period) %>%
  do(metrics = calculate_individual_metrics(.)) %>%
  unnest(metrics)
```

```{r}
# Ensure date column is in proper date format
all_metrics$date <- as.Date(all_metrics$date)
```

```{r}
# Function to safely convert to numeric
safe_as_numeric <- function(x) {
  as.numeric(as.character(x))
}
```

```{r}
# Clean and convert data types
all_metrics <- all_metrics %>%
  mutate(across(c(mean_glucose, variance_glucose, mage, peak_glucose), 
                ~safe_as_numeric(.)))
```

```{r}
# Check for any remaining non-numeric values
non_numeric_check <- all_metrics %>%
  summarise(across(c(mean_glucose, variance_glucose, mage, peak_glucose), 
                   ~sum(is.na(.) & !is.na(.))))

print("Number of non-numeric values after conversion:")
print(non_numeric_check)
```

```{r}
# Remove rows with NA values in key metrics
all_metrics_clean <- all_metrics %>%
  filter(!is.na(mean_glucose) & !is.na(variance_glucose) & !is.na(mage) & !is.na(peak_glucose))
```

```{r}
# Summarize metrics for each individual and period
summary_metrics <- all_metrics %>%
  group_by(id, period) %>%
  summarise(
    mean_glucose = mean(mean_glucose, na.rm = TRUE),
    mean_variance = mean(variance_glucose, na.rm = TRUE),
    mean_mage = mean(mage, na.rm = TRUE),
    max_peak_glucose = max(peak_glucose, na.rm = TRUE),
    days_recorded = n_distinct(date),
    .groups = 'drop'
  )
```

```{r}
# Print summary metrics
print(summary_metrics)

# Save the results to a CSV file
write.csv(summary_metrics, "cgm_summary_metrics.csv", row.names = FALSE)

# Create daily metrics dataset
daily_metrics <- all_metrics %>%
  select(id, period, date, mean_glucose, variance_glucose, mage, peak_glucose)

# Print first few rows of daily metrics
print(head(daily_metrics))

# Save daily metrics to a CSV
write.csv(daily_metrics, "cgm_daily_metrics.csv", row.names = FALSE)
```

################################################################################
################################################################################
################################################################################

Correlation Analysis

```{r}
# Load required libraries if not already loaded
library(tidyverse)
library(corrplot)

# File 2: k5 data
path <- file.choose()
k5_data <- read.csv(path, sep=",", header = TRUE)
```

```{r}
# Subset the columns
k5_data <- data.frame(
  id = k5_data$ID,
  k5 = k5_data$value
)
```

```{r}
# File 3: Clamp data
path <- file.choose()
clamp_data <- read.csv(path, sep=",", header = TRUE)
```

```{r}
# Subset the columns
clamp_data <- data.frame(
  id = clamp_data$ID,
  clamp_M_value = clamp_data$clamp_M_value,
  Matsuda_CIW1 = clamp_data$Matsuda_CI1,
  Matsuda_CIW2 = clamp_data$Matsuda_CI2,
  HOMA.IR_CIW1 = clamp_data$HOMA.IR_CI1,
  HOMA.IR_CIW2 = clamp_data$HOMA.IR_CI2,
  HOMA_beta_CIW1 = clamp_data$HOMA_beta_CI1,
  HOMA_beta_CIW2 = clamp_data$HOMA_beta_CI2,
  IGI_CIW1 = clamp_data$IGI_CI1,
  IGI_CIW2 = clamp_data$IGI_CI2
)
```

Transforming Erdos Clamp so that there is ordering by Period
```{r}
# Step 1: Create CIW1 dataset
data_ciw1 <- data.frame(
  id = clamp_data$id,
  clamp_M_value = clamp_data$clamp_M_value,
  Matsuda = clamp_data$Matsuda_CIW1,
  HOMA_IR = clamp_data$HOMA.IR_CIW1,
  HOMA_beta = clamp_data$HOMA_beta_CIW1,
  IGI = clamp_data$IGI_CIW1 
) %>%
  mutate(period = "CIW1")

# Step 2: Create CIW2 dataset
data_ciw2 <- data.frame(
  id = clamp_data$id,
  clamp_M_value = clamp_data$clamp_M_value,
  Matsuda = clamp_data$Matsuda_CIW2,
  HOMA_IR = clamp_data$HOMA.IR_CIW2,
  HOMA_beta = clamp_data$HOMA_beta_CIW2,
  IGI = clamp_data$IGI_CIW2 
) %>%
  mutate(period = "CIW2")

# Step 3: Combine CIW1 and CIW2 datasets
final_data <- bind_rows(data_ciw1, data_ciw2)

# Step 4: Arrange by id and period
final_data <- final_data %>% arrange(id, period)

# Drop first two rows
clamp_data <- final_data %>% slice(-1, -2)
```

```{r}
# Function to remove prefix from id
remove_prefix <- function(id) {
  ifelse(grepl("^(Ex|PS)", id), sub("^(Ex|PS)", "", id), id)
}
```

```{r}
# Print sample of ids before modification
print("Sample of ids in clamp_data before modification:")
print(head(clamp_data$id))
print("Sample of ids in k5_data before modification:")
print(head(k5_data$id))
print("Sample of ids in summary_metrics before modification:")
print(head(summary_metrics$id))
```
```{r}
# Remove prefixes from both datasets
clamp_data$id <- remove_prefix(clamp_data$id)
k5_data$id <- remove_prefix(k5_data$id)
summary_metrics$id <- remove_prefix(summary_metrics$id)
```

```{r}
# Merge k5 values with summary metrics
k5_summary_metrics_data <- summary_metrics %>%
  left_join(k5_data, by = "id")

# Merge clamp values with summary metrics & k5 values
merged_data <- k5_summary_metrics_data %>%
  left_join(clamp_data, by = "id")
```
```{r}
filtered_data <- merged_data[merged_data$period.x == merged_data$period.y, ]
filtered_data$period.y <- NULL
names(filtered_data)[names(filtered_data) == "period.x"] <- "period"
final_data <- unique(filtered_data)

# Check the result
print(head(final_data))
print(nrow(final_data))
```

```{r}
# Check for any missing values after merging
missing_after_merge <- sum(is.na(merged_data$k5))
print(paste("Number of missing k5 values after merging:", missing_after_merge))

missing_after_merge <- sum(is.na(merged_data$clamp_M_value))
print(paste("Number of missing clamp_M_values after merging:", missing_after_merge))
```

```{r}
# If there are still missing values, let's investigate
if (missing_after_merge > 0) {
  # Identify ids that didn't match
  unmatched_ids <- summary_metrics$id[!summary_metrics$id %in% k5_data$id]
  print("Sample of unmatched ids:")
  print(head(unmatched_ids))
  
  print("Number of unique ids in summary_metrics:")
  print(length(unique(summary_metrics$id)))
  
  print("Number of unique ids in k5_data:")
  print(length(unique(k5_data$id)))
  
  print("Sample of ids in summary_metrics:")
  print(head(summary_metrics$id))
  
  print("Sample of ids in k5_data:")
  print(head(k5_data$id))
}
```

```{r}
final_data[is.na(final_data)] <- NaN
```

```{r}
# Remove rows with missing k5 values for the correlation analysis
merged_data_complete <- final_data %>% 
  filter(!is.na(k5))

merged_data_complete <- merged_data_complete %>%
  filter(!is.na(clamp_M_value))
```

```{r}
threshold <- 0.0000001  # threshold as this is a value for when it deviates
filter_data <- merged_data_complete %>% filter(k5 > threshold)
```

Correlation Analysis
```{r}
# Correlation Analysis
# cor_with_clamp <- cor(filter_data[c("clamp_M_value", "k5", "mean_mage")], use = "pairwise.complete.obs")

cor_with_clamp <- cor(filter_data[c("clamp_M_value", "k5", "mean_mage")], 
                      method = "spearman")

print("Correlations with CLAMP M-value:")
print(cor_with_clamp["clamp_M_value", ])
```

```{r}
# Visualization
ggplot(filter_data, aes(x = k5, y = clamp_M_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "K5 vs CLAMP M-value",
       x = "K5", y = "CLAMP M-value") +
  theme_minimal()
ggsave("k5_vs_clamp.png", width = 8, height = 6)

ggplot(filter_data, aes(x = mean_mage, y = clamp_M_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "MAGE vs CLAMP M-value",
       x = "MAGE", y = "CLAMP M-value") +
  theme_minimal()
ggsave("mage_vs_clamp.png", width = 8, height = 6)
```
Investigate the correlation between CLAMP and all metrics
```{r}
# List all metrics to correlate with CLAMP
metrics <- c("mean_glucose", "mean_variance", "mean_mage", "max_peak_glucose", 
             "k5", "Matsuda", "HOMA_IR", "HOMA_beta", "IGI")

# Ensure all columns are numeric
merged_data <- merged_data %>%
  mutate(across(c(all_of(metrics), clamp_M_value), ~as.numeric(as.character(.))))

# Remove rows only where CLAMP value is NA
merged_data <- merged_data %>%
  filter(!is.na(clamp_M_value))

# Calculate correlations
cor_with_clamp <- cor(merged_data[c("clamp_M_value", all_of(metrics))], 
                      method = "spearman", use = "pairwise.complete.obs")

# Extract correlations with CLAMP
clamp_correlations <- cor_with_clamp["clamp_M_value", metrics]

# Create a data frame of correlations
correlation_df <- data.frame(
  Metric = names(clamp_correlations),
  Correlation = as.numeric(clamp_correlations)
) %>%
  arrange(desc(abs(Correlation)))

print("Correlations with CLAMP M-value:")
print(correlation_df)

# Visualize correlations
ggplot(correlation_df, aes(x = reorder(Metric, abs(Correlation)), y = Correlation)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Correlations with CLAMP M-value", x = "Metric", y = "Correlation") +
  theme_minimal()

ggsave("all_correlations_with_clamp.png", width = 10, height = 6)

# Scatter plots for each metric
for (metric in metrics) {
  p <- ggplot(merged_data, aes_string(x = metric, y = "clamp_M_value")) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(title = paste(metric, "vs CLAMP M-value"),
         x = metric, y = "CLAMP M-value") +
    theme_minimal()
  
  print(p)
  ggsave(paste0(metric, "_vs_clamp.png"), width = 8, height = 6)
}
```
Correlations between K5 and the metrics
```{r}
# List all metrics to correlate with CLAMP
k5_metrics <- c("mean_glucose", "mean_variance", "mean_mage", "max_peak_glucose", 
             "clamp_M_value", "Matsuda", "HOMA_IR", "HOMA_beta", "IGI")

# Ensure all columns are numeric
k5_merged_data <- merged_data %>%
  mutate(across(c(all_of(k5_metrics), k5), ~as.numeric(as.character(.))))

# Remove rows only where CLAMP value is NA
k5_merged_data <- k5_merged_data %>%
  filter(!is.na(k5))

# Calculate correlations
cor_with_k5 <- cor(k5_merged_data[c("k5", all_of(k5_metrics))], 
                      method = "spearman", use = "pairwise.complete.obs")

# Extract correlations with CLAMP
k5_correlations <- cor_with_k5["k5", k5_metrics]

# Create a data frame of correlations
k5_correlation_df <- data.frame(
  Metric = names(k5_correlations),
  Correlation = as.numeric(k5_correlations)
) %>%
  arrange(desc(abs(Correlation)))

print("Correlations with k5:")
print(k5_correlation_df)

# Visualize correlations
ggplot(k5_correlation_df, aes(x = reorder(Metric, abs(Correlation)), y = Correlation)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Correlations with k5", x = "Metric", y = "Correlation") +
  theme_minimal()

ggsave("all_correlations_with_k5.png", width = 10, height = 6)

# Scatter plots for each metric
for (metric in k5_metrics) {
  p <- ggplot(k5_merged_data, aes_string(x = metric, y = "k5")) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(title = paste(metric, "vs k5"),
         x = metric, y = "k5") +
    theme_minimal()
  
  print(p)
  ggsave(paste0(metric, "_vs_k5.png"), width = 8, height = 6)
}
```



################################################################################
################################################################################
################################################################################

```{r}
path <- file.choose()
k5_data <- read.csv(path)
```



